---
title: 'Bellabeat Case Study: Sleep Quality'
author: "Benjamin"
date: "2022-07-30"
output: html_notebook
---

## Introduction

Welcome everyone. This capstone project will mark my very first milestone of DA projects. The objective of the analysis is to provide high-level recommendations to Bellabeat marketing team in order to guide marketing strategy for the company. I will first analyze a open source dataset **"FitBit Fitness Tracker Data"** and reveal some interesting insights of how consumers use smart devices. After that, I will apply these insights into one of the Bellabeat product and provide some data-driven recommendations.

## Background of Bellabeat

Bellabeat is a high-tech company that manufactures health-focused smart products. Collecting data on activity, sleep, stress, and reproductive health has allowed Bellabeat to empower women with knowledge about their own health and habits.

## Business Tasks

First, analyze the activity, intensity, sleep and weight data that smart devices collected and apply the insights onto Bellabeat product. Finally, give high-level recommendations for marketing strategy.

## Data Description - Prepare

In this analysis, I will use FitBit Fitness Tracker Data [link](https://www.kaggle.com/datasets/arashnic/fitbit) to reveal the insights of device users' daily habits. This dataset provides a comprehensive records of 30 fitbit users, with their daily activity, steps, we could even explore users' habits. However, this dataset is inconsistent, and the sample size is limited, we may need to add another data source in order to address these limitations to give a thorough analysis.

```{r Load R packages}
library(tidyverse)
library(tidyr)
library(lubridate)
library(ggplot2)
library(dplyr)
library(janitor)
library(skimr)
library(readr)
library(scales)
```

```{r Loading the data using readr}
weight <- read_csv("Fitabase Data 4.12.16-5.12.16/weightLogInfo_merged.csv")
activity <- read_csv("Fitabase Data 4.12.16-5.12.16/dailyActivity_merged.csv")
step <- read_csv("Fitabase Data 4.12.16-5.12.16/dailySteps_merged.csv")
intensity <- read_csv("Fitabase Data 4.12.16-5.12.16/dailyIntensities_merged.csv")
intensity_hour <- read_csv("Fitabase Data 4.12.16-5.12.16/hourlyIntensities_merged.csv")
calories <- read_csv("Fitabase Data 4.12.16-5.12.16/dailyCalories_merged.csv")
sleep <- read_csv("Fitabase Data 4.12.16-5.12.16/sleepDay_merged.csv")
```

After Loading our data, lets clean up the column names using clean_names() and have a overview of our datasets.

```{r Cleaning the column names}
weight %>% clean_names() %>% head()
activity %>% clean_names() %>% head()
step %>% clean_names() %>% head()
intensity %>% clean_names() %>% head()
calories %>% clean_names() %>% head()
sleep %>% clean_names() %>% head()
```

Next, we will check the completeness of our datasets.

```{r Data Summary_before, echo=TRUE}
Datasets <- c("weight", "activity", "step", "intensity", "calories", "sleep", "intensity_hour")
Unique_ID <- c(n_distinct(weight$Id), n_distinct(activity$Id), n_distinct(step$Id),
               n_distinct(intensity$Id), n_distinct(calories$Id), n_distinct(sleep$Id), n_distinct(intensity_hour$Id))
Duplicates <- c(sum(duplicated(weight)), sum(duplicated(activity)), sum(duplicated(step)), 
                sum(duplicated(intensity)), sum(duplicated(calories)), sum(duplicated(sleep)), sum(duplicated(intensity_hour)))
Missing <- c(sum(is.na(weight)), sum(is.na(activity)), sum(is.na(step)),
             sum(is.na(intensity)), sum(is.na(calories)), sum(is.na(sleep)), sum(is.na(intensity_hour)))

Summary <- data.frame(Datasets, Unique_ID, Duplicates, Missing)

Summary
```

As we can see there are only 8 distinct users in weight data which is not complete. And there are only 24 distinct users recording their sleep data. Although it is not complete, we can still get some interesting insights.Also, there are 65 missing records in *weight* dataset and 3 duplicates in *sleep* dataset.

## Data Manipulation - Process

First, we will remove any duplicated records and missing values. We can observe that those 65 missing records in weight dataset are actually from variable *Fat*. Therefore, we decided to drop the Fat column to prevent data bias. Furthermore, we will remove the 3 duplicates in *sleep* dataset as well.

```{r Remove duplicated and null records}
weight$Fat <- NULL
sleep <- sleep %>% distinct()
```

```{r Data Summary_after}
Duplicates <- c(sum(duplicated(weight)), sum(duplicated(activity)), sum(duplicated(step)), 
                sum(duplicated(intensity)), sum(duplicated(calories)), sum(duplicated(sleep)), sum(duplicated(intensity_hour)))
Missing <- c(sum(is.na(weight)), sum(is.na(activity)), sum(is.na(step)),
             sum(is.na(intensity)), sum(is.na(calories)), sum(is.na(sleep)), sum(is.na(intensity_hour)))

Summary <- data.frame(Datasets, Unique_ID, Duplicates, Missing)
Summary
```

Next, we can observe that the *date* column is now stored as string in R, we want to transform the date variable into date or datetime format in R.

```{r Transforming Date from string to date format}
weight <- weight %>% mutate(Date=parse_date_time(Date, "%m/%d/%Y %I:%M:%S %p"))
activity <- activity %>% mutate(ActivityDate=as.Date(ActivityDate, format="%m/%d/%Y")) %>% rename(Date = ActivityDate)
step <- step %>% mutate(ActivityDay=as.Date(ActivityDay, format="%m/%d/%Y")) %>% rename(Date = ActivityDay)
intensity <- intensity %>% mutate(ActivityDay=as.Date(ActivityDay, format="%m/%d/%Y")) %>% rename(Date = ActivityDay)
calories <- calories %>% mutate(ActivityDay=as.Date(ActivityDay, format="%m/%d/%Y")) %>% rename(Date = ActivityDay)
sleep <- sleep %>% mutate(SleepDay=parse_date_time(SleepDay, "%m/%d/%Y %I:%M:%S %p")) %>% rename(Date = SleepDay)
intensity_hour <- intensity_hour %>% mutate(ActivityHour=parse_date_time(ActivityHour, "%m/%d/%Y %I:%M:%S %p"))
intensity_hour <- intensity_hour %>% mutate(Time=format(ActivityHour, format = "%H:%M"),
                                            Date=format(ActivityHour, format = "%m/%d/%Y"))
```

## Patterns and Insights - Analyse & Share

### Intensity vs. Sleep

```{r Intensity}
intensity_sleep <- merge(intensity,sleep, by=c('Id','Date'))
intensity_sleep <- intensity_sleep %>% mutate(TotalActiveMinutes = VeryActiveMinutes + 
                                                FairlyActiveMinutes + LightlyActiveMinutes + SedentaryMinutes)
## Donut Chart Data
Total <- mean(intensity_sleep$TotalActiveMinutes)
VeryActive <- round(mean(intensity_sleep$VeryActiveMinutes) / Total * 100, digit = 1)
FairlyActive <- round(mean(intensity_sleep$FairlyActiveMinutes) / Total * 100, digit = 1)
LightlyActive <- round(mean(intensity_sleep$LightlyActiveMinutes) / Total * 100, digit = 1)
SedentaryActive <- round(mean(intensity_sleep$SedentaryMinutes) / Total * 100, digit = 1)

donut_cat <- c("Very Active", "Fairly Active", "Lightly Active", "Sedentary Active")
donut_data <- c(VeryActive, FairlyActive, LightlyActive, SedentaryActive)
hsize <- 2

donut <- data_frame(donut_cat, donut_data, hsize)

## Donut Chart of Intensity Data
ggplot(donut, aes(x = hsize, y = donut_data, fill = donut_cat)) +
  geom_col() +
  geom_text(aes(label = donut_data),
             position = position_stack(vjust = 0.5),
            size = 3, color="#352961", show.legend=FALSE) +
  xlab("") +
  ylab("") +
  coord_polar(theta='y', direction=-1) +
  xlim(c(0.2, hsize + 0.5)) +
  scale_fill_manual(values=c("#AD5C6C", "#FBB637", "#E76236", "#C00000")) +
  theme(panel.background = element_rect(fill = "white"),
        panel.grid = element_blank(),
        axis.title = element_blank(),
        axis.ticks = element_blank(),
        axis.text = element_blank()) +
  labs(title = "Average Distribution of Activeness", subtitle = "73% of the users time are in sedentary active",
       fill = "Activeness Category")
```

The above donut chart shows that 73% of the users time are in sedentary active.

```{r Sedentary time vs. Sleep}
ggplot(data = intensity_sleep, mapping=aes(x=SedentaryMinutes, y=TotalMinutesAsleep)) +
  geom_point() +
  geom_smooth() +
  labs(title = "Sedentary time versus sleeping time ", subtitle = "The longer the sedentary time, the shorter the sleeping time",
       x = "Sedentary Time", y = "Sleeping Time", caption = "in minutes")
```

As we can see, there is a small negative correlation between the variables. The data follows the smooth line slanting down from left to right. The longer the people sit, the relatively shorter time they are asleep.

```{r Instensity}
avg_intensity <- intensity_hour %>% 
  group_by(Time) %>% 
  summarize(avg_TotalIntensities = mean(TotalIntensity))

ggplot(data = avg_intensity, aes(x=Time, y=avg_TotalIntensities, fill=avg_TotalIntensities)) +
  geom_bar(stat = "identity") +
  theme(axis.text.x = element_text(angle = 90)) +
  scale_fill_gradient(low = "yellow", high = "red") +
  labs(title = "Average intensity throughout the day", subtitle = "Most users are more active between 8AM to 8PM",
       x = "Time", y = "Average Intensity", fill="Average Intensity") + 
  annotate("text", x=20,y=23, label="5PM to 7PM Most active period", color="black",
           fontface="bold", size=3)
```

We can observed that most users are more active between 8AM to 8PM. While 5PM to 8PM is the peak active period. The graph also shows the average intensity at 3PM is relatively lower than other afternoon time. People might be taking afternoon break or nap during that period of time.

### Sleeping Quality

```{r Sleeping Quality among intensity}
intensity_sleep_df <- intensity_sleep %>% mutate(TotalActiveMinutes=FairlyActiveMinutes + VeryActiveMinutes)

intensity_sleep_df <- intensity_sleep_df %>% group_by(Id) %>% summarize(AvgActiveMinutesPerWeek=mean(TotalActiveMinutes) * 7)

intensity_sleep_df <- intensity_sleep_df %>%
  mutate(Category=case_when(AvgActiveMinutesPerWeek < 150 ~ "Sedentary",
                            AvgActiveMinutesPerWeek >= 150 & AvgActiveMinutesPerWeek <= 300 ~ "Active",
                            AvgActiveMinutesPerWeek > 300 ~ "Highly Active")) %>%
  mutate(Category=factor(Category, levels=c("Sedentary", "Active", "Highly Active"))) %>%
  arrange(Category)

user_category <- intensity_sleep_df %>% select(Id,AvgActiveMinutesPerWeek,Category)

sum_user_category <- user_category %>%
  group_by(Category) %>%
  summarize(Count=n())

hole_size <- 2
ggplot(sum_user_category, aes(x=hole_size, y=Count, fill=Category)) +
  geom_col() +
  coord_polar(theta='y', direction=-1) +
  xlim(c(0.2, hole_size + 0.5)) +
  scale_fill_manual(values=c("#AD5C6C", "#FBB637", "#E76236")) +
  geom_text(aes(label=Count), size=5, position=position_stack(vjust=0.5), color="#352961", show.legend=FALSE) +
  xlab("") + ylab("") + labs(title="Number of Users by Amount of Physical Activity", fill="Category", caption = "There are in total 24 unique users") +
  theme(axis.title = element_blank(),
        axis.ticks = element_blank(),
        axis.text = element_blank(),
        legend.title=element_text(color="#333333"),
        legend.text=element_text(color="#333333"),
        panel.background=element_rect(fill="white", color="white"))
```

```{r Time needed to fall asleep}
intensity_sleep <- intensity_sleep %>% mutate(TimeNeededToSleep = TotalTimeInBed - TotalMinutesAsleep)

df <- intensity_sleep %>%
  group_by(Id) %>% 
  summarize(Avg_TimeNeededToSleep = mean(TimeNeededToSleep)) %>% 
  data_frame()

merged <- merge(df, user_category, by='Id')

ggplot(data = merged, aes(Category,Avg_TimeNeededToSleep)) +
  labs(title="Time Needed To Fall Asleep", subtitle="Sedentary users take longer time to fall asleep", y="Average minutes needed to fall asleep", x="User Category") +
  theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank()) +
  stat_summary(fun=mean, geom="bar", fill=c("#AD5C6C", "#FBB637", "#E76236"))
```

The plot shows us that users who are classified as "Sedentary" take longer time to fall asleep.

## Recommendation - Act

-   Since the longer the people sitting, the shorter time they are asleep or the longer time needed to fall asleep. We suggest using **Bellebeat App** to send reminder to the users, notifying them for sitting or lying for too long and suggesting them to stand or walk or stretch for a while to improve their sleeping qualities. Also, we can suggest users to workout or exercise to improve their sleeping qualities.

-   Most users are more active between 8AM to 8PM, peak during 5PM to 7PM. We suggest to encourage users using **Bellebeat App** to schedule their daily rountine and starting the cycle from 8AM. We can also deploy **Spring** bottle to send out hydration reminders based on users intensity.

-   Besides, we also suggest Bellabeat to educate more users to record their weight and sleep data in order to to keep track of their physical health. Having more complete weight and sleep data, we can further analyse the users health and provide data-driven knowledge and suggestions about their own health and habits.
